
*Practical Machine Learning Course Project*  
*Rodney Gurdal*  
*July 24, 2016*  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

Load libraries:
```{r, warning=FALSE}
library(caret)
library(rpart)
library(rattle)
library(e1071)
library(randomForest)

```


### Get the data, prepare the data  

The "testingdata"" set is set aside. Model exploration and analysis are only performed on the "trainingdata"" set.

```{r}
filename1 <- "pml-training.csv"
fileURL1 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"   
if (!file.exists(filename1)){
  download.file(fileURL1, filename1, method="libcurl")
}  
if (!file.exists("trainingdata")) { 
  trainingdata <- read.csv("pml-training.csv") 
}


filename2 <- "pml-testing.csv"
fileURL2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"   
if (!file.exists(filename2)){
  download.file(fileURL2, filename2, method="libcurl")
}  
if (!file.exists("testingdata")) { 
  testingdata <- read.csv("pml-testing.csv") 
}

```



### Cross-Validation
Because we want to be able to estimate the out-of-sample error, we randomly split the full training data (trainingdata) 
into a smaller training set (trainingdata1) and a validation set (trainingdata2):
```{r}
set.seed(123)
inTrain <- createDataPartition(y=trainingdata$classe, p=0.7, list=F)
trainingdata1 <- trainingdata[inTrain, ]
trainingdata2 <- trainingdata[-inTrain, ]  
```


The original dataset includes 19622 observations of 160 variables. Many of those variables, however, are not needed. 
We now reduce the dataset by removing variables with nearly zero variance, variables
that are almost always NA, and variables that don’t make sense for prediction. After these variables 
are removed from "trainingdata1", we remove the same ones from "trainingdata2".

```{r}
# remove variables with nearly zero variance
nzv <- nearZeroVar(trainingdata1)
trainingdata1 <- trainingdata1[, -nzv]
trainingdata2 <- trainingdata2[, -nzv]

# remove variables that are almost always NA
mostlyNA <- sapply(trainingdata1, function(x) mean(is.na(x))) > 0.95
trainingdata1 <- trainingdata1[, mostlyNA==F]
trainingdata2 <- trainingdata2[, mostlyNA==F]

# remove variables that don't make  sense for prediction (user_name, etc.).
trainingdata1 <- trainingdata1[, -(1:5)]
trainingdata2 <- trainingdata2[, -(1:5)]
```



###Model


We start with a Random Forest model. We  fit the model on trainingdata1, and instruct the “train” function 
to use 3-fold cross-validation to select optimal tuning parameters for the model.

```{r}
fitControl <- trainControl(method="cv", number=3)
RFfit <- train(classe ~ ., data=trainingdata1, method="rf", metric="Kappa", trControl=fitControl)
RFfit$finalModel
```


We'll also try a Gradient Boosting (gbm) model for comparison.
```{r}
GBMfit <- train(classe~., data=trainingdata1, method="gbm", metric="Kappa", trControl=fitControl,verbose=FALSE)
GBMfit$finalModel
```


The models are then compared using the resamples function from the Caret package.
```{r}
rValues <- resamples(list(rf=RFfit,gbm=GBMfit))
summary(rValues)
```



###Plot the results
```{r}
bwplot(rValues,metric="Kappa",main="RandomForest (rf) vs Gradient Boosting (gbm)")
```


We see from the plot that the Random Forest model works better, with a mean accuracy of .996, compared to .9851 for the gbm model. 
We therefore choose the Random Forest model.

###Cross Validation
```{r}
#Training set:
trainingPred <- predict(RFfit, trainingdata1)
confusionMatrix(trainingPred, trainingdata1$classe)

#Validation set:
cvPred <- predict(RFfit, trainingdata2)
confusionMatrix(cvPred, trainingdata2$classe)
```


###Model Testing
```{r}
results <- predict(RFfit,newdata=testingdata)
print(as.data.frame(results))
```



